Time stretching of the GeV emission of GRBs: \\ Fermi-LAT data vs geometrical model

### 2015-12-18 APJ/gammaRaysAPJ.tex
Added comments on curvature effects

Gamma-ray bursts (GRBs) are among the most energetic events in the
Universe. Therefore, the studies of the emission mechanism and
phenomenology of GRBs may provide the new knowledge in particle
physics. Moreover, the GRBs are observed from cosmological distances
and therefore bear an imprint of the late-time evolution of the
Universe. An extensive studies of these explosions led to a number of
interesting results, see~\citet{Vianello:2013ela,Gehrels:2013xd} for a
review. In particular, there are indications that the total energy emitted in gamma-rays during a
burst is similar for the different GRBs within an order
of magnitude~\citep{postnov1999grb,Bloom:2003wy}.
One should keep in mind, though, that the width of the distribution differs for multiple authors, and, in fact, is model dependent. More recent results, for example, suggest that energetics of different bursts might be different~\citep{racusin2009jet, chandra2012radio}. In this paper, however, we explore the possibility that most of the
bursts have similar energetics in their rest frames to simplify analysis. When more precise energy distribution becomes available, the analysis in this paper may be improved. {\bf Also, it is generally accepted that} the
high energy radiation of GRBs follow{\bf s} the predictions of
the synchrotron radiation scenario~\citep{Wang:2013ptaw}, {\bf although there are problems with that model~\citep{Daigne:2010fb} and there are bursts in which additional radiation component clearly appears~\citep{Abdo:2009pg} }. Several
observations are related to the temporal variations of spectra. This
way, the spectral lags were found between different low energy bands {\bf \citep{cheng1995time,Wu:1999ms} }and the very-high-energy radiation was discovered to
be extended relative to the x-ray emission
\citep{Castignani:2014gaa,Lange:2013uh,Vianello:2013ela}. In this paper
we go further and compare the temporal extension of GRB gamma-ray
radiation in the two energy bands $\unit[100]{MeV} < E <
\unit[1]{GeV}$ and $E > \unit[1]{GeV}$ (hereafter, low and high-energy
bands) using the data from the Large Area Telescope (LAT) of the Fermi
satellite~\citep{2009ApJ...697.1071A,Ackermann:2012kna}.

@@

We propose that the observed energy dependence of the time profiles
may be explained without introduction of the new effects. We argue
that the stretching may originate from the effects of the jet geometry
called curvature effects. These effects were explored by multiple
authors {\bf \citep{Meszaros:1997je,Soderberg:2001mk,Nakamura:2001kd,Shen:2005ea,Shenoy:2013cba,granot2002off}}.  These
studies concerned x-ray radiation, and considered the homogeneous
distribution of radiation sources throughout the jet. Our model is
based on the reverse assumption that the highest energy radiators are
concentrated near the axis of the jet and consequently the jet opening
angle depends on energy. The latter is motivated by observations if
all bursts are considered equivalent. In this case the energy
dependence of the jet opening angle resolves the contradiction between
the small fraction of GRBs visible above GeV and the hard
spectrum.

>>>>

Gamma-ray bursts (GRBs) are among the most energetic events in the
Universe. Therefore, the studies of the emission mechanism and
phenomenology of GRBs may provide the new knowledge in particle
physics. Moreover, the GRBs are observed from cosmological distances
and therefore bear an imprint of the late-time evolution of the
Universe. An extensive studies of these explosions led to a number of
interesting results, see~\citet{Vianello:2013ela,Gehrels:2013xd} for a
review. {\bf In particular, there are indications that the total energy emitted in gamma-rays during a
burst is similar for the different GRBs within an order
of magnitude~\citep{postnov1999grb,Bloom:2003wy}.
One should keep in mind, though, that there is no agreement on the width of the distribution, and it is in fact model dependent. More recent results, for example, suggest that energetics of different bursts might be different~\citep{racusin2009jet, chandra2012radio}. In this paper, however, we explore the possibility that most of the
bursts have similar energetics in their rest frames to simplify analysis. When more precise energy distribution becomes available, the analysis in this paper may be improved. {\bf Also, it is generally accepted that} the
high energy radiation of GRBs follow{\bf s} the predictions of
the synchrotron radiation scenario~\citep{Wang:2013ptaw}}, {\bf although there are problems with that model~\citep{Daigne:2010fb} and there are bursts in which additional radiation component clearly appears~\citep{Abdo:2009pg} }. Several
observations are related to the temporal variations of spectra. This
way, the spectral lags were found between different low energy bands {\bf \citep{cheng1995time,Wu:1999ms} }and the very-high-energy radiation was discovered to
be extended relative to the x-ray emission
\citep{Castignani:2014gaa,Lange:2013uh,Vianello:2013ela}. In this paper
we go further and compare the temporal extension of GRB gamma-ray
radiation in the two energy bands $\unit[100]{MeV} < E <
\unit[1]{GeV}$ and $E > \unit[1]{GeV}$ (hereafter, low and high-energy
bands) using the data from the Large Area Telescope (LAT) of the Fermi
satellite~\citep{2009ApJ...697.1071A,Ackermann:2012kna}.

@@

We propose that the observed energy dependence of the time profiles
may be explained without introduction of the new effects. We argue
that the stretching may originate from the effects of the jet geometry
called curvature effects. These effects were explored by multiple
authors {\bf \citep{Meszaros:1997je,Soderberg:2001mk,Nakamura:2001kd,Shen:2005ea,Shenoy:2013cba,granot2002off}}. {\bf These
studies either considered the homogeneous
distribution of radiation sources throughout the jet, or assumed dependence of energy and Lorentz factor on the opening angle. Our model is
based on a similar assumption that the highest energy radiators are
concentrated near the axis of the jet and consequently the jet opening
angle depends on energy. The latter is motivated by observations if
all bursts are considered equivalent. In this case the energy
dependence of the jet opening angle resolves the contradiction between
the small fraction of GRBs visible above GeV and the hard
spectrum. We assume the Lorentz factor to be constant throughout the jet for simplicity.}
